---
title: "Video Game Sales Data Cleaning in R"
author: "Theresa Gyamfi Allotey"
date: "2025-05-15"
output:
  html_notebook: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document outlines the data cleaning process for the `vgsales.csv` dataset, sourced from Kaggle. The dataset contains 16,598 records of video game sales across 11 columns: `Rank`, `Name`, `Platform`, `Year`, `Genre`, `Publisher`, `NA_Sales`, `EU_Sales`, `JP_Sales`, `Other_Sales`, and `Global_Sales`. The goal is to clean the data by addressing missing values, duplicates, and data types, producing `vgsales_cleaned.csv` for downstream analysis or visualization (e.g., in Tableau).

This process showcases data cleaning skills essential for data analytics, including handling missing data, transforming variables, and ensuring data quality, aligned with industry-standard practices.

## Load Required Libraries

Load R packages for data manipulation, summary statistics, and exploratory data analysis (EDA). Install if not already available.

```{r load-libraries, eval=FALSE}
# Install packages if not present
if (!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if (!require(skimr)) install.packages("skimr", repos = "http://cran.us.r-project.org")
if (!require(DataExplorer)) install.packages("DataExplorer", repos = "http://cran.us.r-project.org")

library(tidyverse)     # For data manipulation and visualization
library(skimr)         # For summary statistics
library(DataExplorer)  # For automated EDA
```

## Read the Dataset

Load the dataset, treating "N/A" and empty strings as missing values.

```{r read-data}
df <- read.csv("/kaggle/input/video-game-sales/vgsales.csv", na.strings = c("N/A", ""))
```

## Initial Data Inspection

Inspect the dataset to identify structure, dimensions, and issues like missing values or incorrect data types.

### Check Dimensions

Verify the number of rows and columns.

```{r dim}
dim(df)
```

### Inspect Structure

Examine data types and sample values.

```{r structure}
str(df)
```

### View First Few Rows

Preview the data to confirm content.

```{r head}
head(df)
```

### Summary Statistics

Summarize missing values and distributions using `skimr`.

```{r skim}
skim(df)
```

### Automated EDA

Visualize missing values and distributions with `DataExplorer`.

```{r eda, eval=TRUE}
plot_missing(df)   # Missing data visualization
plot_histogram(df) # Histograms for numeric columns
```

## Data Cleaning Steps

Based on inspection, address:

- Missing `Year` values: Impute with median.
- Unknown publishers: Replace with `NA`, impute with mode.
- Duplicates: Remove exact duplicates.
- Skewed sales: Add `Global_Sales_log` for visualizations.
- Categorical variables: Convert `Platform`, `Genre`, `Publisher` to factors.
- Missing sales: Impute with 0.

### Step 1: Handle Missing Year Values

Impute missing `Year` values with the median to support time-based analysis.

```{r clean-year}
# Check missing Year values
sum(is.na(df$Year))

# Convert Year to numeric
df$Year <- as.numeric(df$Year)

# Impute with median
df$Year[is.na(df$Year)] <- median(df$Year, na.rm = TRUE)

# Verify
sum(is.na(df$Year))
```

### Step 2: Handle Unknown Publishers

Replace "Unknown" in `Publisher` with `NA` and impute with the mode (most frequent publisher).

```{r clean-publisher}
# Check Unknown publishers
sum(df$Publisher == "Unknown", na.rm = TRUE)

# Replace Unknown with NA
df$Publisher[df$Publisher == "Unknown"] <- NA

# Calculate mode publisher
mode_publisher <- names(sort(table(df$Publisher), decreasing = TRUE))[1]

# Impute NA with mode
df$Publisher[is.na(df$Publisher)] <- mode_publisher

# Verify
sum(df$Publisher == "Unknown", na.rm = TRUE)
sum(is.na(df$Publisher))
```

### Step 3: Remove Duplicates

Remove exact duplicates to avoid overcounting.

```{r clean-duplicates}
# Check duplicates
sum(duplicated(df))

# Remove duplicates
df <- df[!duplicated(df), ]

# Verify
sum(duplicated(df))
dim(df)
```

### Step 4: Add Log-Transformed Global Sales

Add `Global_Sales_log` using `log1p` to handle skewness for visualizations.

```{r add-log-sales}
# Add log-transformed Global_Sales
df$Global_Sales_log <- log1p(df$Global_Sales)

# Preview
summary(df$Global_Sales_log)
```

### Step 5: Convert Categorical Columns to Factors

Convert `Platform`, `Genre`, and `Publisher` to factors for grouping in analysis.

```{r convert-factors}
# Convert to factors
df$Platform <- as.factor(df$Platform)
df$Genre <- as.factor(df$Genre)
df$Publisher <- as.factor(df$Publisher)

# Verify
str(df)
```

### Step 6: Handle Missing Sales Values

Impute missing sales values with 0, assuming no sales recorded.

```{r clean-sales}
# Check missing sales
colSums(is.na(df[, c("NA_Sales", "EU_Sales", "JP_Sales", "Other_Sales", "Global_Sales")]))

# Impute with 0
df$NA_Sales[is.na(df$NA_Sales)] <- 0
df$EU_Sales[is.na(df$EU_Sales)] <- 0
df$JP_Sales[is.na(df$JP_Sales)] <- 0
df$Other_Sales[is.na(df$Other_Sales)] <- 0
df$Global_Sales[is.na(df$Global_Sales)] <- 0

# Verify
colSums(is.na(df[, c("NA_Sales", "EU_Sales", "JP_Sales", "Other_Sales", "Global_Sales")]))
```

### Step 7: Save Cleaned Dataset

Save the cleaned dataset as `vgsales_cleaned.csv`.

```{r save-data}
# Save cleaned dataset
write.csv(df, "vgsales_cleaned.csv", row.names = FALSE)

# Confirm file creation
file.exists("vgsales_cleaned.csv")
```

## Final Data Inspection

Verify the cleaned dataset’s structure and completeness.

```{r final-inspection}
# Dimensions
dim(df)

# Structure
str(df)

# Missing values
plot_missing(df)
```